---
title: "Thesis Data Analysis"
author: "Lauren Deitz"
date: '2023-05-10'
output: pdf_document
always_allow_html: true
---

# Maternal Mortality and Abortion Access


## Data Wrangling Steps
This document does not have the data wrangling steps I took, which were in both R and Jupyter notebook:

#### Distances
- The dataset goes back to 2009 and is maintained through today. I restricted to 2010-2020 to keep in line with the population data I have.
- Rolled up to state and year level (rather than county and month) by taking the average distance by state by year.

#### Small Area Health Insurance Estimates
- This data was pulled using the `censusapi` R package and joined with the population/race data below.

#### Population/race
- This data was pulled using the `tidycensus` R package and joined with SAHIE data above.

#### Mortality rate
- I downloaded the mortality datasets by state by year. There are a few states that have suppressed data where fewer than 10 people died of maternal reasons within a year. I filled those suppressed values with 0.
- I downloaded natality (live births) by state by year. This did not result in any suppressed data.
- I used the WHO definition of maternal mortality rate (deaths over 100,000 live births) and calculated using the formula (maternal deaths / live births) * 100,000

#### Overall dataset
- The dataset was joined together in Jupyter notebook.


```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(dplyr)
library(plm)
library(ggplot2)
library(ggthemes)
library(lubridate)
library(networkD3)
library(lmtest)
library(usmap)
library(tidyr)
library(jtools)
library(stargazer)
```


5-way and 3-way score time trend graphs

```{r message=FALSE, warning=FALSE}
dataset <- read.csv("~/Courses/Thesis/thesis_dataset_draft2.csv")

trim_dataset <- dataset %>% 
  mutate(yr = as.Date(ISOdate(YEAR, 1, 1)), yr = format(yr, '%Y'))

# 5-way
grp_data <- trim_dataset %>% 
  group_by(score5way,yr) %>% 
  dplyr::summarise(mn = mean(maternal_mortality_rate))

grp_data$score5way <- factor(grp_data$score5way, 
                             levels=c("Hostile","Restricted","Neutral",
                                      "Supportive","Protective"))

grp_data %>% ggplot(., aes(x=yr,y=mn, color=score5way, group=score5way)) + geom_line() +
  scale_color_manual(values=c("#FF0000","#FF6666","#999999","#33FF99","#009966"), 
                     name="Policy Scores") + theme_clean() + 
  ggtitle("Maternal mortality rate by abortion policy, 5-way score") +
  xlab("Year") + ylab("Maternal mortality rate")
```

```{r message=FALSE, warning=FALSE}
# 3-way

grp_data3 <- trim_dataset %>% 
  group_by(score3way,yr) %>% 
  dplyr::summarise(mn = mean(maternal_mortality_rate))

grp_data3$score3way <- factor(grp_data3$score3way, 
                              levels=c("Hostile","Neutral","Protective"))

grp_data3 %>% ggplot(., aes(x=yr,y=mn, color=score3way, group=score3way)) + geom_line() +
  scale_color_manual(values=c("#FF0000","#999999","#009966"), name="Policy Scores") + 
  theme_clean() + ggtitle("Maternal mortality rate by abortion policy, 3-way score") +
  xlab("Year") + ylab("Maternal mortality rate")
```


Sankey visualization

```{r message=FALSE, warning=FALSE}

#sankey
nodes <- read.csv("~/Courses/Thesis/img/sankey2nodes.csv", fileEncoding="UTF-8-BOM")
links <- read.csv("~/Courses/Thesis/img/sankey2links.csv", fileEncoding="UTF-8-BOM")

links <- links %>% mutate(source = source-1, target=target-1)

library(htmlwidgets)

nodes$group <- as.factor(c("a","b","c","d","e","a","b","c","d","e","a","b","c","d","e"))

links <- links %>% mutate(group = case_when(
  target == 5 ~ "h",
  target == 6 ~ "r", 
  target == 7 ~ "n", 
  target == 8 ~ "s", 
  target == 9 ~ "p",
  target == 10 ~ "h",
  target == 11 ~ "r",
  target == 12 ~ "n",
  target == 13 ~ "s",
  target == 14 ~ "p"
))

my_color <- 'd3.scaleOrdinal() .domain(["a","b","c","d","e", "h", "r", "n", 
"s", "p"]) .range(["#b53226", "#e89d74", "#b3abae", "#ba9bde", "#754ba6","#b53226", 
"#e89d74", "#b3abae", "#ba9bde", "#754ba6"])'

sk <- sankeyNetwork(Links = links, Nodes = nodes, Source = "source", Target = "target", 
                    Value = "value", NodeID = "name", units = "State", fontSize=15, 
                    nodeWidth = 30, colourScale=my_color, NodeGroup="group", 
                    LinkGroup="group")

sk

```

Summary stats

```{r message=FALSE, warning=FALSE}

dataset <- dataset %>% 
  dplyr::rename("black_non_hispanic" = "pp",
                "distance_to_clinic" = "distance_origintodest")

vars <- c("distance_to_clinic", "insurance_rate", "black_non_hispanic", 
          "maternal_mortality_rate", "ScoreNum") 

colnames(dataset)

d.vars <- dataset %>% 
  dplyr::select(vars) %>% 
  dplyr::select("Distance to Clinic" = distance_to_clinic, 
                "Insurance rate" = insurance_rate, 
                "% Black non-Hispanic" = black_non_hispanic, 
                "Maternal mortality rate" = maternal_mortality_rate, 
                "Policy Score" = ScoreNum)

stargazer(d.vars, type="text")

```

Histograms

```{r message=FALSE, warning=FALSE}
firstD <- function(var, group, df){
  bad <- (missing(group) & !missing(df))
  if (bad) stop("if df is specified then group must also be specified")
  
  fD <- function(j){ c(NA, diff(j)) }
  
  var.is.alone <- missing(group) & missing(df)
  
  if (var.is.alone) {
    return(fD(var))
  }
  if (missing(df)){
    V <- var
    G <- group
  }
  else{
    V <- df[, deparse(substitute(var))]
    G <- df[, deparse(substitute(group))]
  }
  
  G <- list(G)
  D.var <- by(V, G, fD)
  unlist(D.var)
}

dataset.fdiff <- firstD(ScoreNum, origin_state, dataset)

dataset.fdiff.na <- dataset.fdiff[!is.na(dataset.fdiff)]

mean(dataset.fdiff.na)

hist(dataset.fdiff, main="State policy changes, n = 550", 
     ylim=c(0,450), xlab="Policy score changes by year, n = 550", 
     labels=TRUE, breaks=-3:3)

dataset.decade <- dataset %>% filter(YEAR == 2010 | YEAR == 2020)

dataset.decade.score.fdiff <- firstD(ScoreNum, origin_state, dataset.decade)

dataset.decade.score.fdiff.na <- 
  dataset.decade.score.fdiff[!is.na(dataset.decade.score.fdiff)]

mean(dataset.decade.score.fdiff.na)

hist(dataset.decade.score.fdiff.na, 
     main="Policy score changes between 2010-2020, n = 550", ylim=c(0,18), 
     xlab="Insurance rate change", labels=TRUE)


```


```{r message=FALSE, warning=FALSE}
dataset.fdiff.insur <- firstD(insurance_rate, origin_state, dataset)

dataset.fdiff.insur.na <- dataset.fdiff.insur[!is.na(dataset.fdiff.insur)]

mean(dataset.fdiff.insur.na)

hist(dataset.fdiff.insur, main="Insurance rate changes by year, n = 550", 
     ylim=c(0,225), xlab="Insurance rate change", labels=TRUE)

dataset.decade <- dataset %>% filter(YEAR == 2010 | YEAR == 2020)

dataset.decade.fdiff <- firstD(insurance_rate, origin_state, dataset.decade)

datasef.decade.fdiff.na <- dataset.decade.fdiff[!is.na(dataset.decade.fdiff)]

mean(datasef.decade.fdiff.na)


hist(dataset.decade.fdiff, main="Insurance rate changes between 2010-2020, n = 550", 
     ylim=c(0,18), xlab="Insurance rate change", labels=TRUE)

```

```{r message=FALSE, warning=FALSE}
dataset.fdiff.pp <- firstD(black_non_hispanic, origin_state, dataset)

dataset.fdiff.pp.na <- dataset.fdiff.pp[!is.na(dataset.fdiff.pp)]

mean(dataset.fdiff.pp.na)

hist(dataset.fdiff.pp.na, main="Black Non-Hispanic changes by year, n = 550", 
     ylim=c(0,259), xlab="% Black Non-Hispanic change", labels=TRUE)

dataset.decade.pp <- dataset %>% filter(YEAR == 2010 | YEAR == 2020)

dataset.decade.pp.fdiff <- firstD(black_non_hispanic, origin_state, dataset.decade.pp)

dataset.decade.pp.fdiff.na <- dataset.decade.pp.fdiff[!is.na(dataset.decade.pp.fdiff)]

mean(dataset.decade.pp.fdiff.na)

hist(dataset.decade.pp.fdiff, 
     main="Black Non-Hispanic changes between 2010-2020, n = 550", ylim=c(0,30), 
     xlab="% Black Non-Hispanic change", labels=TRUE)

```

Map visualization

```{r fig.height=8, fig.width=12}
#map
mapdf <- stack(dataset.decade.score.fdiff.na)

mapdf <- mapdf %>% mutate(state = substr(ind, 1, 2)) %>% dplyr::select(state, values)

plot_usmap(data=mapdf, color="white") + 
  scale_fill_gradient2(low="#b53226", mid="#b3abae", high="#754ba6",
                       name="Policy Score Change 2010-2020") + 
  labs("Policy Score Changes in the United States") +  theme(legend.position="right")
```
OLS models

```{r message=FALSE, warning=FALSE}
ols1 <- lm(maternal_mortality_rate ~ ScoreNum + insurance_rate + distance_to_clinic + 
             as.factor(YEAR), dataset)

## simple OLS ##
summary(ols1)

# clustered standard errors
coeftest(ols1, cluster = origin_state)

```

Fixed effects models

```{r message=FALSE, warning=FALSE}

fx1 = plm(maternal_mortality_rate ~ ScoreNum + as.factor(YEAR),
            index = c('origin_state','YEAR'),
            model = 'within',
            data = dataset)

summary(fx1)

fx2 = plm(maternal_mortality_rate ~ ScoreNum + insurance_rate + distance_to_clinic + 
            as.factor(YEAR),
            index = c('origin_state','YEAR'),
            model = 'within',
            data = dataset)

summary(fx2)

```

```{r message=FALSE, warning=FALSE}

jtools::export_summs(ols1, coeftest(ols1, cluster = origin_state), fx1, fx2, 
                     model.names = c("OLS", "Clustered SE", "Fixed Effects 1", "Fixed Effects 2"), 
                     coefs = c("Policy score" = "ScoreNum",
                               "Distance to facility" = "distance_to_clinic",
                               "State insurance rate" = "insurance_rate",
                               "2011" = "as.factor(YEAR)2011",
                               "2012" = "as.factor(YEAR)2012",
                               "2013" = "as.factor(YEAR)2013",
                               "2014" = "as.factor(YEAR)2014",
                               "2015" = "as.factor(YEAR)2015",
                               "2016" = "as.factor(YEAR)2016",
                               "2017" = "as.factor(YEAR)2017",
                               "2018" = "as.factor(YEAR)2018",
                               "2019" = "as.factor(YEAR)2019",
                               "2020" = "as.factor(YEAR)2020"
                               ),
                     to.file = "Word",
                     file.name = "modelsummary.docx")

```